
```{r , include=FALSE}
#tutorial::go_interactive(greedy = TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```

# ADVANCED BASE PLOTTING TECHNIQUES

## Combining Different Graph Types
We can overlay different types of plots, such as combining bar charts with line graphs, in R. Here's how this can be done  along with customizing axes and scales:

### Overlay different types of plots

```{r}
# Sample data
x <- 1:10
y1 <- x^2
y2 <- 2 * x
y3 <- x^1.5

# Create a line plot
plot(x, y1, type = "l", col = "blue", main = "Combining Different Graph Types", xlab = "X", ylab = "Y")

# Add points to the line plot
points(x, y1, col = "blue", pch = 16)

# Add a bar plot
barplot(y2, col = "red", add = TRUE)

# Add a second line plot
lines(x, y3, col = "green")

# Add a legend
legend("topleft", legend = c("Trend 1", "Trend 2", "Trend 3"), col = c("blue", "red", "green"), lty = c(1, NA, 1), pch = c(16, NA, NA))

```


### Axes and scales customization to accomodate combined plots

```{r}
# Set custom axis limits
plot(x, y1, type = "l", col = "blue", main = "Combining Different Graph Types", xlab = "X", ylab = "Y", ylim = c(0, max(y1, y2, y3)))

# Add points to the line plot
points(x, y1, col = "blue", pch = 16)

# Add a bar plot
barplot(y2, col = "red", add = TRUE, ylim = c(0, max(y1, y2, y3)))

# Add a second line plot
lines(x, y3, col = "green")

# Add a legend
legend("topleft", legend = c("Trend 1", "Trend 2", "Trend 3"), col = c("blue", "red", "green"), lty = c(1, NA, 1), pch = c(16, NA, NA))

```



## Interactive Graphics in Base R

### Introduction to using the locator() function for interactive plotting

Base R provides basic functionality for creating interactive plots using the locator() function. Here's how you can use locator() for interactive plotting and create dynamic plots that respond to user input or updates in data:


```{r}
# Sample data
x <- 1:10
y <- x^2

# Plot data
plot(x, y, type = "l", main = "Interactive Plotting with locator()")

# Add points interactively
points(locator(5), col = "red", pch = 16)

```


### Creating dynamic plots that respond to user input or updates in data

```{r}
# Dynamic plot that responds to user input
plot_dynamic <- function() {
  # Sample data
  x <- 1:10
  y <- x^2
  
  # Plot data
  plot(x, y, type = "l", main = "Dynamic Plot with User Input")
  
  # Add points interactively
  points(locator(5), col = "red", pch = 16)
}

# Call the function to create dynamic plot
plot_dynamic()

```


