

```{r , include=FALSE}
#tutorial::go_interactive(greedy = TRUE)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
```


# STATISTICAL GRAPHS IN R

## Introduction

The two main objectives of this chapter are:

- Learn to use R’s base graphics for creating statistical plots.
- Understand the application of statistical plots in real-world data analysis.


## Essential Statistical Plots:
Statistical plots are essential for exploring and understanding the distribution of data. In this chapter, we will cover how to create the following essential statistical plots in R and understand the insights they provide.

### How to create box plots, violin plots, and QQ plots.

**i) Box Plots:**

```{r, height=500}
# Creating a box plot
boxplot(Sepal.Length ~ Species, data = iris, 
        main = "Box Plot of Sepal Length by Species",
        xlab = "Species", ylab = "Sepal Length")

```


**ii) Violin Plots:**

```{r , height=500}
# Creating a violin plot
library(ggplot2)
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
  geom_violin(fill = "lightblue") +
  labs(title = "Violin Plot of Sepal Length by Species",
       x = "Species", y = "Sepal Length")

```


**iii) QQ Plots:**

```{r, height=500}
# Creating a QQ plot
qqnorm(iris$Sepal.Length)
qqline(iris$Sepal.Length)

```


### Understanding the data insights these plots provide.

**i) Box Plots:** Box plots provide a graphical summary of the distribution of a dataset. They display the median, quartiles, and potential outliers.

**ii) Violin Plots:** Violin plots are similar to box plots but provide a more detailed summary of the distribution. They combine a box plot with a kernel density plot.

**iii) QQ Plots (Quantile-Quantile Plots):** QQ plots are used to assess if a dataset follows a particular distribution. They compare the quantiles of the dataset against the quantiles of a theoretical distribution.


These statistical plots are invaluable tools for exploring and understanding the distribution of data.

## Analyzing Distributions with Histograms and Density Plots:
Histograms provide a visual representation of the distribution of numerical data. They group data into bins and display the frequency of observations in each bin.Density plots estimate the probability density function of the underlying data distribution. They provide a smooth representation of the distribution.

### Techniques for creating and customizing histograms and density plots and best practices for displaying distribution characteristics.

**Creating and Customizing Histograms:**
```{r, height=500}
# Creating a histogram
hist(iris$Sepal.Length, 
     main = "Histogram of Sepal Length",
     xlab = "Sepal Length",
     ylab = "Frequency",
     col = "lightblue")

# Adding a density plot
lines(density(iris$Sepal.Length), col = "red")

```


**Creating and Customizing Density Plots:**

```{r, height=500}
# Creating a density plot
plot(density(iris$Sepal.Length), 
     main = "Density Plot of Sepal Length",
     xlab = "Sepal Length",
     col = "blue")

# Adding a histogram
hist(iris$Sepal.Length, 
     col = rgb(0,0,1,0.2), 
     add = TRUE, 
     breaks = 20)

```



## Scatter Plots and Correlation Analysis:

Scatter plots are useful for visualizing the relationship between two continuous variables.

### Using scatter plots to visualize relationships between variables.

Scatter plots visually represent the relationship between two continuous variables. They help to identify patterns, trends, and correlations between variables.

```{r, height=500}
# Creating a scatter plot
plot(iris$Sepal.Length, iris$Petal.Length,
     main = "Scatter Plot of Sepal Length vs Petal Length",
     xlab = "Sepal Length", ylab = "Petal Length",
     col = "blue")

```



### Introduction to adding regression lines and confidence intervals.

- Regression lines provide a summary of the relationship between variables.

- Confidence intervals help assess the uncertainty of the estimated regression line.


```{r, height=500}
# Creating a scatter plot
plot(iris$Sepal.Length, iris$Petal.Length,
     main = "Scatter Plot and Regression Line ",
     xlab = "Sepal Length", ylab = "Petal Length",
     col = "blue")

# Fit linear regression model
model <- lm(Petal.Length ~ Sepal.Length, data = iris)

# Adding regression line
abline(model, col = "red")

# Adding regression equation
eq <- paste("y =", round(coef(model)[1], 2), "+", round(coef(model)[2], 2), "x")
text(6.5, 2.5, eq, pos = 4)

# Adding confidence intervals
conf_interval <- predict(model, interval = "confidence")
lines(iris$Sepal.Length, conf_interval[, "lwr"], col = "green", lty = 2)
lines(iris$Sepal.Length, conf_interval[, "upr"], col = "green", lty = 2)
# Add legend
legend("topleft", legend = c("Data Points", "Regression Line", "Confidence Intervals"),
       col = c("blue", "red", "green"), lty = c(NA, 1, 2), lwd = c(NA, 1, 1),
       pch = c(1, NA, NA), bty = "o" # Enclose the legend in a box
       )

```



## Practical Examples: Exercises


**Exercise 15**

Create a box plot that compares the distribution of Sepal Length across the three species in the iris dataset.


```{r, ex="exercise_num7.1", type="sample-code", height=500}
# Your code
```


**Exercise 16**

Generate a random health dataset of 100 people (male and female) from a normal distribution with the three variables: Age,BMI and Gender. The mean of Age should be 50 with a stand deviation of 10. The mean of BMI should be 25 with a standard deviation of 4. Gender should have only two values; "Male" and "Female". Set random seed to a value of 123 and name the data frame as health_dataset. Create a scatter plot that examines the relationship between Age and BMI in the simulated health dataset. Include a linear regression line with the linear regression equation and the value of Adjusted R-Squared to highlight the trend.


```{r, ex="exercise_num7.2", type="sample-code", height=500}
# Your code
```


